The Weakly Nonlinear Saturation of the Magnetorotational Instability

### 2014-12-15 paper/mri_paper.tex
hmri description

\beq
\mathbf{B} = B_0 \hat{z},
\eeq

@@

We conduct a full study of the MRI parameter space, focussed on the behavior with changing magnetic Prandtl number ($\mathrm{Pm} \equiv \frac{\mathrm{Rm}}{\mathrm{Re}} \equiv \frac{\nu}{\eta}$). In protoplanetary disks, as well as in liquid metal MRI experiments, the magnetic Prandtl number can be very small ($\mathrm{Pm} \sim 10^{-6}$). The scale separations caused by these large ratios of diffusivity to viscosity render the small Prandtl number regime inaccessible to simulations. This analytic formulation allows us to explore the behavior down to magnetic Prandtl numbers of [TBD!].

@@

The present configuration, that of magnetized Taylor Couette flow with a purely axial magnetic field, represents the initial conditions for the standard MRI. The helical MRI is a related, but separate, MHD instability, and was first discovered by Hollerbach and R\"udiger (\citeyear{Hollerbach:2005tr}). It arises in the presence of combined axial and azimuthal fields. To seed helical MRI in our current set-up, we need only add an azimuthal component to our initial background magnetic field. This will be an avenue of future study, and will yield an analytic theory relevant to the above-mentioned PROMISE experiment.

>>>>

\beq
\label{backgroundfield1}
\mathbf{B} = B_0 \hat{z},
\eeq

@@

We conduct a full study of the MRI parameter space, focussed on the behavior with changing magnetic Prandtl number ($\mathrm{Pm} \equiv \frac{\mathrm{Rm}}{\mathrm{Re}} \equiv \frac{\nu}{\eta}$). In protoplanetary disks, as well as in liquid metal MRI experiments, the magnetic Prandtl number can be very small ($\mathrm{Pm} \sim 10^{-6}$). The scale separations caused by these large ratios of diffusivity to viscosity render the small Prandtl number regime inaccessible to simulations. This analytic formulation allows us to explore the behavior down to magnetic Prandtl numbers of $10^{-6}$ or lower.

\section{Helical MRI}

The configuration investigated thus far, that of magnetized Taylor Couette flow with a purely axial magnetic field, constitutes the standard MRI. The helical MRI (HMRI) is a related, but separate, MHD instability, first identified by Hollerbach and R\"udiger (\citeyear{Hollerbach:2005tr}). It arises in the presence of combined axial and azimuthal fields.

We initialize the helical MRI by adding a constant background toroidal field to Equation \ref{backgroundfield1}, such that our initial magnetic field structure becomes

\beq
\mathbf{B} = B_0 \hat{z} + \xi B_0 \hat{y},
\eeq

where $\xi$ is a dimensionless parameter representing the relative strength of the axial field and the vertical field.

This addition leaves Equations \ref{eqset2} - \ref{eqset4} unchanged, but Equation \ref{eqset1} gains a term to become

\begin{multline}
\label{eqset1_hmri}
\partial_t \nabla^2 \Psi \, = \, \frac{2}{\beta} B_0 \partial_z \nabla^2 A \, + \, 2 \partial_z u_{y} \, - \xi B_0 \partial_x \partial_z B_{y}\\
- \, J\left(\Psi, \nabla^2 \Psi\right) \, + \, \frac{2}{\beta}J\left(A, \nabla^2 A \right) \, + \, \frac{1}{\reye}\nabla^4 \Psi,
\end{multline}

in other words, the $\Psi$ equation picks up a linear $B_y$ term.

Following the same prescription for weakly nonlinear tuning, multiscale expansion, and perturbation expansion, we find the equations for the HMRI up to order $epsilon^3$.

To $\mathcal{O}(\epsilon)$, we find

\mathcal{L} \mathbf{V_1}

@@

%The present configuration, that of magnetized Taylor Couette flow with a purely axial magnetic field, represents the initial conditions for the standard MRI. The helical MRI is a related, but separate, MHD instability, and was first discovered by Hollerbach and R\"udiger (\citeyear{Hollerbach:2005tr}). It arises in the presence of combined axial and azimuthal fields. To seed helical MRI in our current set-up, we need only add an azimuthal component to our initial background magnetic field. This will be an avenue of future study, and will yield an analytic theory relevant to the above-mentioned PROMISE experiment.
